---
title: "Homework 7"
format: html
---

**Due:** Friday, March 20, 11:59pm on Canvas

**Instructions:** For mathematical work, you may choose to either hand-write your work and submit a PDF scan, or type your work using LaTeX and submit the resulting PDF. For computational work, include all code and R output necessary to answer the questions, and submit as an html or pdf on Canvas.

See the course website for a [homework template](https://sta711-s26.github.io/homework/hw_template.tex) file and [instructions](https://sta711-s26.github.io/homework/latex_instructions/) on getting started with LaTeX and Overleaf.

## Variance stabilizing transformation

Let $Y_1,...,Y_n \overset{iid}{\sim} IG(\mu, 1)$ be an iid sample from the *inverse Gaussian* distribution, with moment generating function

$$M_Y(t) = \exp \left\lbrace \frac{1}{\mu} \left(1 - \sqrt{1 - 2 \mu^2 t}\right) \right\rbrace$$

::: {.callout-note icon=false}

## Question 1

1. In this question we will look at the distribution of $\overline{Y}$.
    (a) Find the limiting distribution of $\sqrt{n}(\overline{Y} - \mu)$.
    (b) Find a variance stabilizing transformation $g$, so that the asymptotic variance of $\sqrt{n}(g(\overline{Y}) - g(\mu))$ does not depend on $\mu$.

:::
    
    
## Sandwich variance -- Poisson regression

Suppose that we have iid observations $(\mathbf{x}_i, Y_i)$ from some distribution $g$. We *assume* that the data come from the Poisson regression model:

$$
\begin{aligned}
Y_i | \mathbf{x}_i &\sim Poisson(\mu_i) \\
\log(\mu_i) &= \mathbf{x}_i^T \boldsymbol{\beta}
\end{aligned}
$$
However, our assumption could be wrong! We therefore decide to use the sandwich variance estimator. Let $\widehat{\boldsymbol{\beta}}$ be our estimator obtained by maximizing the Poisson likelihood (which may or may not be the true likelihood). Then, recall that we have shown that

$$
\sqrt{n}(\widehat{\boldsymbol{\beta}} - \boldsymbol{\beta}^0) \overset{d}{\to} N(\boldsymbol{0}, \mathbf{S}(\boldsymbol{\beta}^0))
$$

where $\mathbf{S}(\boldsymbol{\beta}^0)$ is the sandwich variance.

::: {.callout-note icon=false}

## Question 2

2. In this question, we will explore the asymptotic distribution of $\widehat{\boldsymbol{\beta}}$.
    (a) Explain what $\boldsymbol{\beta}^0$ represents in this setting, in terms of some expectation with respect to the true distribution $g$.
    (b) Analogous to the class example with linear regression, find an expression for $\mathbf{S}(\boldsymbol{\beta}^0)$ in terms of expectations (with respect to $g$) involving $\mathbf{x}_i$, $Y_i$, and $\mu_i$.
    (c) Give the plug-in estimator for $\mathbf{S}(\boldsymbol{\beta}^0)$ in terms of the observed $\mathbf{x}_i$, $Y_i$, and $\widehat{\boldsymbol{\beta}}$.

:::

### Simulation: the effect of model mis-specification

Now let's use some simulations to explore the behavior of $\widehat{\boldsymbol{\beta}}$ when the model is mis-specified. For these simulations, suppose that we observe iid data $(\mathbf{x}_i, Y_i)$ where

$$
\begin{aligned}
Y_i | \mathbf{x}_i &\sim NB(\mu_i, \theta) \\
\log(\mu_i) &= \mathbf{x}_i^T \boldsymbol{\beta}
\end{aligned}
$$

where $NB(\mu_i, \theta)$ is the *negative binomial* distribution with mean $\mathbb{E}[Y_i|\mathbf{x}_i] = \mu_i$ and variance $Var(Y_i | \mathbf{x}_i) = \mu_i + \frac{\mu_i^2}{\theta}$.

In this model, we can see that we have correctly specified the **systematic** component of the model (the log mean is a linear function of the explanatory variables). However, we have incorrectly specified the random component. As we saw in HW 1, there is a close connection between the negative binomial and the Poisson. For large $\theta$, $NB(\mu_i, \theta) \approx Poisson(\mu_i)$. But for small $\theta$, the negative binomial has much more variability than the Poisson. So, for small $\theta$, our Poisson model will substantially underestimate the true variability in the data.

The code below simulates $n = 200$ observations from the model

$$X_i \overset{iid}{\sim} N(0, 1) \hspace{1cm} Y_i|X_i \sim NB(\mu_i = \exp\{1 + 0.5X_i \}, \theta = 0.5)$$
We then fit a Poisson regression model to estimate the coefficients $\beta_0 = 1$ and $\beta_1 = 0.5$. Again note that, by construction, our *systematic* component of the model is correctly specified (the mean function), but our *random* component (the Poisson distribution) is *not* correct.

```{r, eval=F}
set.seed(523494)

n <- 200
theta <- 0.5
x <- rnorm(n)
mu <- exp(1 + 0.5*x)
y <- rnbinom(n, size = theta, mu = mu)

m1 <- glm(y ~ x, family = poisson)
```

::: {.callout-note icon=false}

## Question 3

3. Let $\widehat{\beta}_{1, \text{Pois}}$ be the estimate of $\beta_1$ from the Poisson fit. In this question, we will use simulations to explore the behavior of $\widehat{\beta}_{1, \text{Pois}}$.
    (a) Using the expressions found in question 2, compute the robust (sandwich) variance estimator on the simulated data: $\widehat{Var}(\widehat{\boldsymbol{\beta}}_{\text{Pois}}) = \frac{1}{n}\widehat{\mathbf{S}}$
    (b) The *naive* variance estimator assumes that the Poisson regression model is *correct*, and can be calculated with `vcov(m1)`. Compare the naive and robust variance estimators for the simulated data.
    (c) Repeat the simulation many times. At each iteration, store $\widehat{\beta}_{1, \text{Pois}}$, and both the robust and naive estimators of $Var(\widehat{\beta}_{1, \text{Pois}})$.
    (d) Using your results from part (c), verify that $Bias(\widehat{\beta}_{1, \text{Pois}}) \approx 0$. (This is because the systematic component of the model is correctly specified).
    (e) Using your results from part (c), empirically estimate $Var(\widehat{\beta}_{1, \text{Pois}})$, and compare to the average estimates from the naive and robust approaches. Which method does better at estimating $Var(\widehat{\beta}_{1, \text{Pois}})$?
    (f) Repeat parts (c) and (e) with different values of $\theta$ in the Negative Binomial. How does the relative performance of the naive and robust variance estimators change as $\theta$ increases?
    
:::

```{r, include=F, eval=F}
poisson_sandwich <- function(mod){
  X <- model.matrix(mod)
  W1 <- diag(mod$fitted.values)
  W2 <- diag((mod$y - mod$fitted.values)^2)
  
  solve(t(X) %*% W1 %*% X) %*% t(X) %*% W2 %*% X %*% solve(t(X) %*% W1 %*% X)
}

n <- 200
#theta <- 0.5
theta <- 10

nsim <- 5000
beta1_hat <- rep(NA, nsim)
beta1_var_naive <- rep(NA, nsim)
beta1_var_robust <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n)
  mu <- exp(1 + 0.5*x)
  y <- rnbinom(n, size = theta, mu = mu)
  
  m1 <- glm(y ~ x, family = poisson)
  beta1_hat[i] <- coef(m1)[2]
  beta1_var_naive[i] <- vcov(m1)[2,2]
  beta1_var_robust[i] <- poisson_sandwich(m1)[2,2]
}

hist(beta1_hat)
mean(beta1_hat)
var(beta1_hat)
mean(beta1_var_naive)
mean(beta1_var_robust)
```

### Comparison with the MLE

Though the Poisson model was mis-specified, $\widehat{\boldsymbol{\beta}}_{\text{Pois}}$ is still consistent for the true mean function because the systematic component was correct. Furthermore, using the sandwich variance allowed us to correctly estimate $Var(\widehat{\boldsymbol{\beta}}_{\text{Pois}})$.

The next question to ask is: how does the Poisson estimator compare to the *true* MLE (i.e., the MLE obtained by correctly assuming a Negative Binomial)?

The estimator $\widehat{\boldsymbol{\beta}}_{\text{NB}}$ obtained by maximum likelihood estimation with the Negative Binomial distribution can be calculated in R using the `glm.nb` function from the `MASS` package. Here is an example:

```{r, eval=F}
set.seed(523494)

n <- 200
theta <- 0.5
x <- rnorm(n)
mu <- exp(1 + 0.5*x)
y <- rnbinom(n, size = theta, mu = mu)

m1 <- glm(y ~ x, family = poisson)
m2 <- MASS::glm.nb(y ~ x)
```

::: {.callout-note icon=false}

## Question 4

4. Let's compare $\widehat{\boldsymbol{\beta}}_{\text{Pois}}$ to $\widehat{\boldsymbol{\beta}}_{\text{NB}}$ in simulations.
    (a) Repeat the provided simulation code many times. For each iteration, store the slope estimates $\widehat{\beta}_{1, \text{Pois}}$ and $\widehat{\beta}_{1, \text{NB}}$ from the Poisson and Negative Binomial models, respectively.
    (b) Using the results from part (a), compare the variances of $\widehat{\beta}_{1, \text{Pois}}$ and $\widehat{\beta}_{1, \text{NB}}$. Which estimator should we prefer?
    (c) With reference to asymptotic properties of MLEs, explain why the results in part (b) should be expected.
    (d) Repeat (a) and (b) with different values of $\theta$. How does the relative performance of $\widehat{\beta}_{1, \text{Pois}}$ and $\widehat{\beta}_{1, \text{NB}}$ change as $\theta$ increases? Why?

:::


```{r, include=F, eval=F}
n <- 200
#theta <- 0.5
theta <- 10

nsim <- 1000
beta1_hat_pois <- rep(NA, nsim)
beta1_hat_nb <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n)
  mu <- exp(1 + 0.5*x)
  y <- rnbinom(n, size = theta, mu = mu)
  
  m1 <- glm(y ~ x, family = poisson)
  m2 <- MASS::glm.nb(y ~ x)
  
  beta1_hat_pois[i] <- coef(m1)[2]
  beta1_hat_nb[i] <- coef(m2)[2]
}

var(beta1_hat_pois)
var(beta1_hat_nb)
```

### Moral

The point here is that, if parts of our model are correct (or close enough to correct), then maximum likelihood estimation with a mis-specified model can still be useful for parameter estimation. Furthermore, the sandwich variance can approximate the true variance of our parameter estimates even though the model is mis-specified.

However, we also know that maximum likelihood estimation is asymptotically efficient -- maximum likelihood estimation with a correctly-specified model will beat incorrect model + sandwich variance.

## M-estimation: sample mean and variance

Suppose we have iid data $Y_1,...,Y_n$ from some distribution $g$ with finite mean and variance $\mathbb{E}_g[Y_i] = \mu$ and $Var_g(Y_i) = \sigma^2$. The estimated mean and variance from the sample moments are

$$\widehat{\mu} = \frac{1}{n} \sum \limits_{i=1}^n Y_i = \overline{Y} \hspace{1cm} \widehat{\sigma}^2 = \frac{1}{n} \sum \limits_{i=1}^n (Y_i - \overline{Y})^2$$
Let $\boldsymbol{\theta}_0 = (\mu, \sigma^2)^T$ be the true mean and variance, and $\widehat{\boldsymbol{\theta}} = (\widehat{\mu}, \widehat{\sigma}^2)^T$.

::: {.callout-note icon=false}

## Question 5

5. The purpose of this question is to use our M-estimation results to analyze the asymptotic behavior of $\widehat{\boldsymbol{\theta}}$.
    (a) Find a function $\psi(Y_i, \boldsymbol{\theta})$ such that $\widehat{\boldsymbol{\theta}}$ is the solution to
    $$\sum \limits_{i=1}^n \psi(Y_i, \boldsymbol{\theta}) = \boldsymbol{0}$$
    (*Hint*: $\psi$ will be a vector-valued function, since $\widehat{\boldsymbol{\theta}}$ needs to solve a system of two equations with two unknowns!)
    (b) Show that the sandwich variance is given by
    $$\mathbf{S}(\boldsymbol{\theta}_0) = \begin{pmatrix} \sigma^2 & \mathbb{E}_g[(Y_i - \mu)^3] \\ \mathbb{E}_g[(Y_i - \mu)^3] & \mathbb{E}_g[(Y_i - \mu)^4] - \sigma^2 \end{pmatrix}$$
    
:::

::: {.callout-note icon=false}

## Question 6
    
6. **If** $Y_i \overset{iid}{\sim} N(\mu, \sigma^2)$, then $\widehat{\boldsymbol{\theta}}$ is also the maximum likelihood estimator, and previously you have shown that the Fisher information is 
$$\mathcal{I}_1(\boldsymbol{\theta}_0) = \begin{pmatrix} \frac{1}{\sigma^2} & 0 \\ 0 & \frac{1}{2\sigma^4} \end{pmatrix}$$
Show that, if $Y_i \overset{iid}{\sim} N(\mu, \sigma^2)$, then your sandwich variance from question 5 is equal to
$$\mathbf{S}(\boldsymbol{\theta}_0) = \begin{pmatrix} \sigma^2 & 0 \\ 0 & 2 \sigma^4 \end{pmatrix} = \mathcal{I}_1^{-1}(\boldsymbol{\theta}_0)$$
(This is what we should expect, since we can't have two different asymptotic variances for the same estimator!) You may use, without proof, the fact that if $Y_i \sim N(\mu, \sigma^2)$, then for any $q \in \mathbb{N}$,
$$
\begin{aligned}
\mathbb{E}[(Y_i - \mu)^{2q + 1}] &= 0 \\
\mathbb{E}[(Y_i - \mu)^{2q}] &= \sigma^{2q}(2q - 1)!! = \sigma^{2q}(2q - 1)(2q - 3) \cdots 3 \cdot 1 \\
\end{aligned}
$$
:::

## Empirical cdfs, quantiles, and M-estimation

So far, our class has focused mostly on *parametric* estimation: we write down a parametric model for the data, and fit the model using techniques like maximum likelihood estimation. However, estimation need not be parametric -- there are a wide variety of *nonparametric* approaches to estimating distributions and statistical functionals. Here we will explore estimation of the cdf and quantiles for an arbitrary distribution, without requiring any parametric assumptions on the distribution.

### Empirical cdf

Let $Y_1,...,Y_n$ be an iid sample from some distribution with cdf $F$. No assumptions are made about the form of $F$. The natural estimator of $F$ is called the **empirical cdf**, $F_n$, and is defined as:

$$F_n(t) = \frac{1}{n} \sum \limits_{i=1}^n \mathbb{I}\{ Y_i \leq t \}$$

::: {.callout-note icon=false}

## Question 7

7. Show that, for all $t$, $F_n(t) \overset{p}{\to} F(t)$. (In fact, this result can be strengthened to *uniform* convergence, a result of the *Glivenko-Cantelli* theorem, but we will not prove that here).

:::

::: {.callout-note icon=false}

## Question 8

8. For a fixed $t$, find the limiting distribution of $\sqrt{n}(F_n(t) - F(t))$.

:::


### Quantiles and M-estimation

For $p \in (0, 1)$, the $p$th quantile of $F$ is the value $\theta = F^{-1}(p)$. That is, $F(\theta) = p$. For example, the **median** is the quantile with $p = 0.5$. 

We estimate quantiles of $F$ with the quantiles of the empirical cdf $F_n$. In particular, the sample $p$th quantile is $\widehat{\theta} = F_n^{-1}(p)$. Written another way, $F_n(\widehat{\theta}) \approx p$. And to be a bit more formal:
$$|F_n(\widehat{\theta}) - p| \leq \frac{1}{n}$$
Now, we can write this like an M-estimator! The sample quantile $\widehat{\theta}$ satisfies

$$
\sum \limits_{i=1}^n \psi(Y_i, \widehat{\theta}) = c_n
$${#eq-quantile_m}
where
$$\psi(Y_i, \widehat{\theta}) = \mathbb{I}\{Y_i \leq \widehat{\theta}\} - p$$
and 
$$|c_n| = n|F_n(\widehat{\theta}) - p| \leq 1$$

@eq-quantile_m looks **very similar** to our definition of an M-estimator. However, we have two small problems to deal with before we can apply M-estimator results. First, the right-hand side of @eq-quantile_m is not 0; it is $c_n$. Second, the function $\psi$ has a jump discontinuity at $Y_i$, which causes problems for differentiation. Fortunately, both of these issues can be accounted for in our theory of M-estimation, as described in the following small extension of the results from class.

**Extension:** Suppose $Y_1,...,Y_n$ are iid from some distribution $F$, and $\widehat{\boldsymbol{\theta}}$ is an estimator which satisfies
$$\sum \limits_{i=1}^n \psi(Y_i, \widehat{\boldsymbol{\theta}}) = \boldsymbol{c}_n$$
where $\boldsymbol{c}_n/\sqrt{n} \overset{p}{\to} \boldsymbol{0}$ as $n \to \infty$. Furthermore, suppose that $\psi$ is not necessarily continuous. Then,
$$\sqrt{n}(\widehat{\boldsymbol{\theta}} - \boldsymbol{\theta}_0) \overset{d}{\to} N(\boldsymbol{0}, \mathbf{S}(\boldsymbol{\theta}_0))$$
where
$$\mathbf{S}(\boldsymbol{\theta}_0) = \mathbf{A}^{-1} \mathbf{B} \mathbf{A}^{-1}$$
and
$$\mathbf{A} = -\frac{\partial}{\partial \boldsymbol{\theta}} \mathbb{E}_F[\psi(Y_i, \boldsymbol{\theta})] \ \Biggr\lvert_{\boldsymbol{\theta} = \boldsymbol{\theta}_0}$$
$$\mathbf{B} = \mathbb{E}_F[\psi(Y_i, \boldsymbol{\theta}_0)\psi(Y_i, \boldsymbol{\theta}_0)^T]$$
That is, our matrix $\mathbf{B}$ stays the same as before, but our matrix $\mathbf{A}$ is slightly different: we have switched the order of integration and differentiation. In other words, $\mathbf{A}$ is calculated by taking the derivative of $\mathbb{E}_F[\psi(Y_i, \boldsymbol{\theta})]$ with respect to $\boldsymbol{\theta}$, and evaluating the derivative at $\boldsymbol{\theta}_0$.

Let's apply these results to get an asymptotic distribution for sample quantiles.

::: {.callout-note icon=false}

## Question 9

9. Let $Y_1,...,Y_n \overset{iid}{\sim} F$, for some continuous $F$ with pdf $f$, and let $\theta_0$ be the $p$th quantile of $F$, and $\widehat{\theta}$ the sample $p$th quantile. Using the M-estimation framework and extension above, show that
$$\sqrt{n}(\widehat{\theta} - \theta_0) \overset{d}{\to} N\left(0, \frac{p(1-p)}{f^2(\theta_0)} \right)$$

:::

### Example: sample median

Now let's use the result of question 9 to examine the performance of the sample median for estimating the center of several symmetric distributions, and calculate the asymptotic relative efficiency of the sample mean and median.

::: {.callout-note icon=false}

## Question 10

10. For each of the following distributions, let $Y_1,...,Y_n$ be an iid sample from the specified distribution. Each distribution is symmetric, and has both mean and median $\mu$. Let $\overline{Y}$ be the sample mean, and $M_n$ the sample median. For each distribution, calculate $ARE(\overline{Y}, M_n)$ for estimating $\mu$. You may use the provided mean, median, and variance of each distribution without proof.
    (a) $Y_1,...,Y_n \overset{iid}{\sim} N(\mu, \sigma^2)$
    (b) $Y_1,...,Y_n \overset{iid}{\sim} Logistic(\mu, s)$, with pdf
    $$f(y) = \dfrac{e^{-(y-\mu)/s}}{s(1 + e^{-(y-\mu)/s})^2}$$
    The mean and median are both $\mu$, and the variance is $\frac{s^2 \pi^2}{3}$
    (c) $Y_1,...,Y_n \overset{iid}{\sim} Laplace(\mu, b)$, with pdf
    $$f(y) = \frac{1}{2b} \exp\left\lbrace -\frac{1}{b}|y - u| \right\rbrace$$
    The mean and median are both $\mu$, and the variance is $2b^2$.
    
:::










    
    