---
title: "M-estimation"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Motivating example

Votes for Pat Buchanan vs. votes for George Bush in Florida counties, 2000 election:

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=3, fig.height=2}
library(Stat2Data)
library(tidyverse)
data("PalmBeach")

PalmBeach |>
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), size=0.5) +
  geom_smooth(aes(y = Buchanan), se=F, method="lm", lwd=0.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw()
```

**Question:** What do you notice about this relationship? Anything concerning?

## Motivating example

Comparing model with and without Palm Beach:

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=3.5, fig.height=3}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach |>
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) |>
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), size=0.5) +
  geom_line(aes(y = pred_1), color = "blue", lwd=0.5) +
  geom_line(aes(y = pred_2), color = "red", lty=2, lwd=0.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw()
```

## Robust linear regression

**Model:** $Y_i = \mathbf{x}_i^T \boldsymbol{\beta} + \varepsilon_i$

**OLS:** $\widehat{\boldsymbol{\beta}}$ minimizes 
$$\sum \limits_{i=1}^n (Y_i - \mathbf{x}_i^T \boldsymbol{\beta})^2$$

**Robust regression:** $\widehat{\boldsymbol{\beta}}$ minimizes 
$$\sum \limits_{i=1}^n \rho(Y_i - \mathbf{x}_i^T \boldsymbol{\beta})$$

One choice of $\rho$ is the **Huber loss function**:

\vspace{3cm}

## Huber loss function

$$\rho_k(u) = \begin{cases} \frac{1}{2}u^2 & |u| \leq k \\ k|u| - \frac{1}{2}k^2 & \text{else} \end{cases}$$

```{r, echo=F, message=F, warning=F, fig.width=4, fig.height=3.5}
huber <- function(u, gamma){
  ifelse(abs(u) <= gamma, 0.5*u^2, gamma*abs(u) - 0.5*gamma^2)
}

x <- seq(-2, 2, 0.01)
plot(x, huber(x, 1), type="l", ylab = "Huber loss", 
     xaxt = "n", xlab = "", yaxt="n")
abline(v = c(-1, 1), lty=2)
```

\vspace{3cm}

## Robust linear regression


```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=3.5, fig.height=3}
library(MASS)

lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))
mod3 <- rlm(Buchanan ~ Bush, data = PalmBeach)

PalmBeach |>
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach),
         pred_3 = predict(mod3)) |>
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), size=0.5) +
  geom_line(aes(y = pred_1), color = "blue", lwd=0.5) +
  geom_line(aes(y = pred_2), color = "red", lty=2, lwd=0.5) +
  geom_line(aes(y = pred_3), color = "orange", lty=4, lwd=0.6) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw()
```

## Robust linear regression

**Model:** $Y_i = \mathbf{x}_i^T \boldsymbol{\beta} + \varepsilon_i$

$\widehat{\boldsymbol{\beta}}$ minimizes $\sum \limits_{i=1}^n \rho_k(Y_i - \mathbf{x}_i^T \boldsymbol{\beta})$

\vspace{6cm}

## M-estimation

So far, a lot of our estimators look something like this: $\widehat{\boldsymbol{\theta}}$ solves
$$\sum \limits_{i=1}^n \psi(Y_i, \boldsymbol{\theta}) = \boldsymbol{0}$$
**Sample mean:**

\bigskip

**MLE:**

\bigskip

**Linear regression:**

\bigskip

**Logistic regression:**

\bigskip

**Robust regression:**

\bigskip

Can we say something **in general** about estimators of this type?

## M-estimation

Let $Y_1,...,Y_n$ be iid samples from some distribution $g$. ($Y_i$ could be a scalar or vector random variable). Let $\psi(Y, \boldsymbol{\theta})$ be some suitably smooth function, which depends on a parameter $\boldsymbol{\theta} \in \mathbb{R}^d$.

\vspace{6cm}

## Example: robust linear regression

$\psi(Y_i, \mathbf{x}_i, \boldsymbol{\beta}) = \rho'_k(Y_i - \mathbf{x}_i^T \boldsymbol{\beta}) \mathbf{x}_i$
$$\rho_k(u) = \begin{cases} \frac{1}{2}u^2 & |u| \leq k \\ k|u| - \frac{1}{2}k^2 & \text{else} \end{cases} \hspace{0.5cm} \rho'_k(u) = \begin{cases} u & |u| \leq k \\ k \ \text{sign}(u) & \text{else} \end{cases}$$
\vspace{5cm}


