---
title: "Multiple testing considerations"
author: "Ciaran Evans"
output: beamer_presentation
---

## Some methods to control FWER

Suppose we test $m$ null hypotheses $H_{0,1},...,H_{0,m}$. Let $p_i$ be the corresponding p-value for test $i$.

**Bonferroni method:** Reject $H_{0,i}$ if $p_i < \alpha/m$

\bigskip

**Holm method:**

* Order the p-values $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$
* Let $i^* = \min \left\lbrace i: p_{(i)} > \frac{\alpha}{m - i + 1} \right\rbrace$
* Reject $H_{0, (i)}$ for all $i < i^*$

\vspace{3cm}

## Ordered p-value plots

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=3.5}
library(tidyverse)
library(gridExtra)
set.seed(234)

m <- 100
m0 <- 0.5*m
mu1 <- 0.43
n <- 100

p_vals <- rep(NA, m)
for(i in 1:m){
  if(i <= m0){
    x <- rnorm(n)
  } else {
    x <- rnorm(n, mean=mu1)
  }
  
  p_vals[i] <- 2*pnorm(abs(sqrt(n)*mean(x)), lower.tail=F)
}

p1 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values")

p2 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  geom_hline(yintercept = 0.05) +
  geom_hline(yintercept = 0.05/m, color = "orange") +
  geom_function(fun = function(x) 0.05/(m - x + 1), color = "blue") +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values") +
  scale_y_continuous(limits = c(0, 0.06))

grid.arrange(p1, p2, ncol=2)
```

## Ordered p-value plots

Decreasing $m_0$:

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=3.5}
m <- 100
m0 <- 0.1*m
mu1 <- 0.43
n <- 100

p_vals <- rep(NA, m)
for(i in 1:m){
  if(i <= m0){
    x <- rnorm(n)
  } else {
    x <- rnorm(n, mean=mu1)
  }
  
  p_vals[i] <- 2*pnorm(abs(sqrt(n)*mean(x)), lower.tail=F)
}

p1 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values")

p2 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  geom_hline(yintercept = 0.05) +
  geom_hline(yintercept = 0.05/m, color = "orange") +
  geom_function(fun = function(x) 0.05/(m - x + 1), color = "blue") +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values") +
  scale_y_continuous(limits = c(0, 0.06))

grid.arrange(p1, p2, ncol=2)
```

## Ordered p-value plots

Decreasing $m_0$:

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=3.5}
m <- 100
m0 <- 0.01*m
mu1 <- 0.43
n <- 100

p_vals <- rep(NA, m)
for(i in 1:m){
  if(i <= m0){
    x <- rnorm(n)
  } else {
    x <- rnorm(n, mean=mu1)
  }
  
  p_vals[i] <- 2*pnorm(abs(sqrt(n)*mean(x)), lower.tail=F)
}

p1 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values")

p2 <- data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  geom_hline(yintercept = 0.05) +
  geom_hline(yintercept = 0.05/m, color = "orange") +
  geom_function(fun = function(x) 0.05/(m - x + 1), color = "blue") +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values") +
  scale_y_continuous(limits = c(0, 0.06))

grid.arrange(p1, p2, ncol=2)
```

## Outcomes for multiple hypothesis tests


## False discovery rate

Suppose we test $m$ hypotheses, $m_0$ of which are truly null. Let $V$ denote the number of type I errors, and $R$ the total number of rejections. 
$$FWER = P(V > 0) \hspace{1cm} FDR = \mathbb{E}[FDP]$$

\vspace{6cm}


## The Benjamini-Hochberg procedure

* Order the p-values $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$
* Let $i^* = \max \left\lbrace i: p_{(i)} < \frac{i\alpha}{m} \right\rbrace$
* Reject $H_{0, (i)}$ for all $i \leq i^*$

```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3}
set.seed(17)
m <- 50
m0 <- 0.3*m
mu1 <- 0.25
n <- 100

p_vals <- rep(NA, m)
for(i in 1:m){
  if(i <= m0){
    x <- rnorm(n)
  } else {
    x <- rnorm(n, mean=mu1)
  }
  
  p_vals[i] <- 2*pnorm(abs(sqrt(n)*mean(x)), lower.tail=F)
}

data.frame(index = 1:m, pval = sort(p_vals)) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  geom_hline(yintercept = 0.05) +
  geom_hline(yintercept = 0.05/m, color = "orange") +
  geom_abline(slope = 0.05/(m+1), intercept = 0.05/m, 
              color="green") +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values") +
  scale_y_continuous(limits = c(0, 0.1))
```

## The Benjamini-Hochberg procedure

* Order the p-values $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$
* Let $i^* = \max \left\lbrace i: p_{(i)} < \frac{i\alpha}{m} \right\rbrace$
* Reject $H_{0, (i)}$ for all $i \leq i^*$


```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3}
data.frame(index = 1:30, pval = sort(p_vals)[1:30]) |>
  ggplot() +
  geom_point(aes(x = index, y = pval), size=0.7) +
  geom_hline(yintercept = 0.05) +
  geom_hline(yintercept = 0.05/m, color = "orange") +
  geom_abline(slope = 0.05/(m+1), intercept = 0.05/m, 
              color="green") +
  theme_bw() +
  labs(x = "Index", y = "Ordered p-values") +
  scale_y_continuous(limits = c(0, 0.025))
```

## The Benjamini-Hochberg procedure: intuition


