---
title: "Beginning hypothesis testing"
author: "Ciaran Evans"
output: beamer_presentation
---


## Course status so far

So far:

* How can we estimate parameters / fit a model?
    * MLE, method of moments, M-estimation, and applications to regression
* What makes a good estimator?
    * MSE, unbiased estimators, CRLB, consistency, asymptotic relative efficiency
* What is the large-sample behavior of our estimators?
    * WLLN, CLT, delta method, asymptotic normality of the MLE, asymptotic normality of M-estimators
    
\bigskip
    
Next: how can we use our estimates for inference?

* Hypothesis tests using Normal distributions
* General hypothesis testing framework
* Other methods for testing hypotheses (Neyman-Pearson, LRT)
* Confidence intervals

## Hypothesis tests for a population mean

Let $X_1, X_2,...,X_n$ be an iid sample from a population with unknown mean $\mu$, and known variance $\sigma^2$. We want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$

**Test statistic:**

\vspace{5cm}

## Hypothesis tests for a population mean

Let $X_1, X_2,..., X_n$ be an iid sample from a population with mean $\mu$ and variance $\sigma^2$. We want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$

**Test statistic:** $Z_n = \dfrac{\overline{X}_n -\mu_0}{\sigma/\sqrt{n}}$

What if $\sigma$ is unknown?

\vspace{5cm}


## Hypothesis tests for a population mean

Let $X_1, X_2,..., X_n$ be an iid sample from a population with mean $\mu$ and variance $\sigma^2$. We want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$

**Test statistic:** $Z_n = \dfrac{\overline{X}_n -\mu_0}{\sigma/\sqrt{n}}$

**Rejecting:** Should we reject $H_0$ when $Z_n$ is close to 0, or when $Z_n$ is far away from 0?

\vspace{5cm}


## Hypothesis tests for a population mean

Let $X_1, X_2,...$ be an iid sample from a population with mean $\mu$ and variance $\sigma^2$. We want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$

**Test statistic:** $Z_n = \dfrac{\overline{X}_n -\mu_0}{\sigma/\sqrt{n}}$

**Rejecting:** Reject $H_0$ when $|Z_n| > z_{\alpha/2}$

**p-value:** How do we calculate a p-value?

\vspace{5cm}

## Wald test for one parameter

## Wald test for one parameter: examples

* Population mean: $H_0: \mu = \mu_0$

\vspace{1.5cm}

* Population proportion: $H_0: p= p_0$

\vspace{1.5cm}

* Regression coefficient: $H_0: \beta_j = 0$

\vspace{2cm}

## Testing multiple parameters

Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

Researchers want to know if there is any relationship between white blood cell count or platelet count, and the probability a patient has dengue. 

**Question:** What hypotheses should they test?

\vspace{2cm}

## Testing multiple parameters

```{r include=F}
library(knitr)


dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

```{r, echo=F}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)$coefficients |> round(digits=3)
```

$H_0: \beta_1 = \beta_2 =0$

Can the researchers test their hypotheses using this output?

\vspace{4cm}

## Wald tests for multiple parameters

For the dengue example:

$$\widehat{\boldsymbol{\beta}} = \begin{pmatrix} \widehat{\beta}_0 \\ \widehat{\beta}_1 \\ \widehat{\beta}_2 \end{pmatrix} \hspace{3cm}$$

```{r}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, family = binomial)
coef(m1)
vcov(m1)
```


## Wald tests for multiple parameters

For the dengue example: $\widehat{\boldsymbol{\beta}} = \begin{pmatrix} \widehat{\beta}_0 \\ \widehat{\beta}_1 \\ \widehat{\beta}_2 \end{pmatrix}$

We want to test: $\begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix} = \begin{pmatrix} 0 \\ 0 \end{pmatrix}$

\vspace{5cm}

## Wald tests for multiple parameters

$$H_0: \mathbf{C} \boldsymbol{\beta} = \gamma_0$$
For the dengue example:

```{r}
C <- matrix(c(0, 1, 0,
              0, 0, 1), nrow=2, byrow=T)

C
C %*% coef(m1)
```

## Wald tests for multiple parameters

For the dengue example:

```{r}
C
vcov(m1)
C %*% vcov(m1) %*% t(C)
```



## Wald tests for multiple parameters

* $H_0: \mathbf{C} \boldsymbol{\beta} = \boldsymbol{\gamma}_0$

* Look at $\mathbf{C} \widehat{\boldsymbol{\beta}}$

\vspace{5cm}

## Test statistic and p-value

* $H_0: {\bf C} \boldsymbol{\beta} = \boldsymbol{\gamma}_0$

* ${\bf C} \widehat{\boldsymbol{\beta}} \approx N({\bf C} \boldsymbol{\beta}, \ {\bf C} \mathcal{I}^{-1}(\boldsymbol{\beta}){\bf C}^T)$

* Want to turn ${\bf C} \widehat{\boldsymbol{\beta}}$ into a scalar test statistic

\vspace{5cm}

## Example

$H_0: {\bf C} \boldsymbol{\beta} = \boldsymbol{\gamma}_0$

\vspace{-0.5cm}

$$({\bf C} \widehat{\boldsymbol{\beta}} - \boldsymbol{\gamma}_0)^T ({\bf C} \mathcal{I}^{-1}(\boldsymbol{\beta}) {\bf C}^T)^{-1} ({\bf C} \widehat{\boldsymbol{\beta}} - \boldsymbol{\gamma}_0) \approx \chi^2_q \hspace{1cm} \text{under } H_0$$

```{r, eval=F}
C %*% coef(m1)
```

```{r, echo=F}
c(C %*% coef(m1))
```



```{r}
test_stat <- t(C %*% coef(m1)) %*% 
  solve(C %*% vcov(m1) %*% t(C)) %*%
  (C %*% coef(m1))
test_stat
pchisq(test_stat, 2, lower.tail=F)
```
