---
title: "Interim Analysis"
author: "Ciaran Evans"
output: beamer_presentation
---

## Motivation: Stopping early?

Observe data $X_1,...,X_n$ over $n$ months, with single $X_i$ collected each month. Assume that $X_1,...,X_n \overset{iid}{\sim} N(\mu, 1)$

Want to test $H_0: \mu = 0$ vs. $H_A: \mu > 0$. Let $Z_k = \frac{1}{\sqrt{k}} \sum \limits_{i=1}^k X_i$.

**Decision rule:** Reject $H_0$ if $Z_k > c_1$ or $Z_n > c_2$.

**Bonferroni approach:**

\vspace{2cm}

**General union bound:**

\vspace{3cm}

## Exact probability of a type I error

## Pocock approach

$$P(Z_{k} > c_1 \cup Z_{n} > c_2) = P(Z_{k} > c_1) + P(Z_{k} < c_1, Z_{n} > c_2)$$

**Pocock approach:** 

\vspace{5cm}

## O'Brien-Fleming (OBF) approach

**Motivation:** Put most of $\alpha$ into the final analysis, so significance level of final analysis is close to that of a single test procedure

**Setup:** Observe $X_1,...,X_n$. Perform $R$ total analyses: interim analyses at times $k_1, k_2,...,k_{R-1} < n$, and final analysis at time $n$. 

\vspace{6cm}

## Example

$P(Z_k > c_1 \cup Z_n > c_2 | \mu = 0)=$
$$\begin{aligned}
(1 - \Phi(c_1)) + \int \limits_{-\infty}^{c_1} \left(1 - \Phi\left(\frac{1}{\sqrt{n - k}} (\sqrt{n} c_2 - \sqrt{k} t) \right) \right) \phi(t) dt
\end{aligned}$$

**Example:** $R = 2$ (so $k = n/2$), and $\alpha = 0.05$

\vspace{6cm}

## Drawbacks

There are several drawbacks to the Pocock and OBF approaches:

1. Have to prespecify number of analyses, and when you will conduct them.

2. Published Pocock/OBF bounds assume evenly-spaced tests. (Can calculate your own boundaries for non-evenly spaced tests, but that isn't as easy as looking up existing boundaries)

\vspace{2cm}

## $\alpha$ spending

Suppose we wish to do $R = 5$ total analyses, and we want an overall type I error rate of $\alpha = 0.05$. OBF and Pocock bounds specify the amount of type I error "budget" that we use for each analysis:

```{r, echo=F, message=F, warning=F, fig.width=7, fig.height=3.5}
library(tidyverse)
library(patchwork)

t_vals <- c(0, 0.2, 0.4, 0.6, 0.8, 1)
a1 <- 2 - 2*pnorm(qnorm(0.975)/sqrt(t_vals))
obf_spent <- diff(a1)

a2 <- 0.05*log(1 + (exp(1) - 1)*t_vals)
pocock_spent <- diff(a2)

p1 <- data.frame(time = rep(t_vals[-1], 2), 
           alpha = c(obf_spent, pocock_spent),
           method = rep(c("OBF", "Pocock"), each=5)) |>
  ggplot(aes(x = time, y = alpha, color = method, shape=method)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(x = "Time", y = "Type I error spent per analysis")

p2 <- data.frame(time = rep(t_vals, 2), 
           alpha = c(a1, a2),
           method = rep(c("OBF", "Pocock"), each=6)) |>
  ggplot(aes(x = time, y = alpha, color = method, shape=method)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(x = "Time", y = "Cumulative type I error spent")

p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
```

\vspace{2cm}

## $\alpha$ spending functions

* Non-decreasing function $f(t) : [0, 1] \to [0, \alpha]$ with $f(0) = 0$ and $f(1) = \alpha$, which gives **total** type I error spent by time $t$
* Conduct analyses at times $0 < t_1 < t_2 < \cdots t_R = 1$
* Portion of $\alpha$ spent on analysis $t_i$ is $f(t_i) - f(t_{i-1})$


```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3}

t_vals <- seq(0, 1, 0.01)
a1 <- 2 - 2*pnorm(qnorm(0.975)/sqrt(t_vals))

a2 <- 0.05*log(1 + (exp(1) - 1)*t_vals)

data.frame(time = rep(t_vals, 2), 
           alpha = c(a1, a2),
           method = rep(c("OBF", "Pocock"), each=length(t_vals))) |>
  ggplot(aes(x = time, y = alpha, color = method, lty=method)) +
  geom_line() +
  theme_bw() +
  labs(x = "Time", y = "Cumulative type I error spent")
```

\vspace{2cm}

## $\alpha$ spending functions

**Advantage:** Don't have to specify number of interim analyses, or when the analyses will be performed, ahead of time

```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3}

t_vals <- seq(0, 1, 0.01)
a1 <- 2 - 2*pnorm(qnorm(0.975)/sqrt(t_vals))

a2 <- 0.05*log(1 + (exp(1) - 1)*t_vals)

data.frame(time = rep(t_vals, 2), 
           alpha = c(a1, a2),
           method = rep(c("OBF", "Pocock"), each=length(t_vals))) |>
  ggplot(aes(x = time, y = alpha, color = method, lty=method)) +
  geom_line() +
  theme_bw() +
  labs(x = "Time", y = "Cumulative type I error spent")
```

\vspace{2cm}
