---
title: "Class activity"
format: html
editor: source
---

# Part I: Hypothesis testing with the Titanic data

In this class activity, we will test hypotheses about quantities of interest in the titanic data. Let $Y_i$ denote whether passenger $i$ survived. We will use the model

$$Y_i|\mathbf{x}_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 Age_i + \beta_3 SecondClass_i + \beta_4 ThirdClass_i$$

The Titanic data can be imported into R with

```{r, eval=F}
titanic <- read.csv("https://sta711-s26.github.io/homework/Titanic.csv")
```

And we can fit the logistic regression model:

```{r, eval=F}
m1 <- glm(Survived ~ Sex + Age + as.factor(Pclass), data = titanic, family = binomial)
```



## Questions

1. Use a Wald test to test 

$$H_0: \beta_3 = \beta_4 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_3, \beta_4 \neq 0$$

Recall that the variance-covariance matrix can be extracted with the `vcov(...)` function.

## Questions: Wald test for a contrast

The null hypothesis $H_0: \beta_3 = \beta_4 = 0$ asks whether there is any relationship between passenger class and survival, after accounting for Sex and Age. But now suppose you want to ask a different question: is there a difference in survival between second and third class passengers? 

Neither $\beta_3$ nor $\beta_4$ directly addresses this question. Instead, we want to compare $\beta_3$ and $\beta_4$ to *each other*; our hypotheses are

$$H_0: \beta_3 - \beta_4 = 0 \hspace{1cm} H_A: \beta_3 - \beta_4 \neq 0$$

How should we test these hypotheses? Once again, we can use a Wald test!


2. Find a matrix ${\bf C}$ such that ${\bf C}\beta = \beta_3 - \beta_4$.

The difference $\beta_3 - \beta_4$ is called a *contrast*, and is another example of a linear transformation of $\beta$.


3. Calculate the Wald test statistic and p-value for $H_0: \beta_3 - \beta_4 = 0$.

# Part II: Testing parameters jointly vs. separately

Consider the logistic regression model

$$Y_i|\mathbf{x}_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2}$$

and suppose we wish to test the hypotheses

$$H_0: \beta_1 = \beta_2 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$$

Or, written another way,

$$H_0: \begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix} = \boldsymbol{0} \hspace{1cm} H_A: \begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix} \neq \boldsymbol{0}$$

Our Wald test for the hypotheses

$$H_0: {\bf C}\beta = \gamma_0 \hspace{1cm} H_A: {\bf C}\beta \neq \gamma_0$$

uses the test statistic

$$W = ({\bf C} \widehat{\beta} - \gamma_0)^T({\bf C} \widehat{\mathcal{I}}^{-1}(\beta) {\bf C}^T)^{-1} ({\bf C} \widehat{\beta} - \gamma_0)$$
Under $H_0$, we have that $W \approx \chi^2_q$, where $q$ is the length of ${\bf C}\beta$. The p-value is

$$p = P(\chi^2_q > W)$$

and we reject $H_0$ if $p < \alpha$ for a given significance level $\alpha$.

### Can we instead test the coefficients *separately*?

This tests both $\beta_1$ and $\beta_2$ *jointly*. But suppose we decided to instead test the null hypothesis $H_0: \beta_1 = \beta_2 = 0$ by *separately* testing the two hypotheses $H_{0,1}: \beta_1 = 0$ and $H_{0,2}: \beta_2 = 0$. Our procedure would be:

* Calculate $Z_1 = \dfrac{\widehat{\beta}_1 - 0}{SE_{\widehat{\beta}_1}}$ and $Z_2 = \dfrac{\widehat{\beta}_2 - 0}{SE_{\widehat{\beta}_2}}$
* Calculate a p-value for each test statistic: $p_1 = 2\Phi(-|Z_1|)$ and $p_2 = 2\Phi(-|Z_2|)$
* Reject $H_{0}: \beta_1 = \beta_2 = 0$ if $\text{min}\{p_1, p_2\} < \alpha$

Is this a good idea? The purpose of the following questions is to explore this question.

### Power and type I error

To help answer whether it is a good idea to test our coefficients separately, we need a definition of "good". Recall from earlier stats courses the concepts of *type I error* and *power*.

Broadly (we will formalize this a bit later):

$$\text{type I error rate} = P(\text{reject } H_0 | H_0 \text{ true})$$

$$\text{power} = P(\text{reject } H_0 | H_0 \text{ false})$$

The usual approach to hypothesis testing in statistics is to try to *maximize* power, subject to control of the type I error rate. That is, we want $\text{type I error rate} < \alpha$ for some chosen value $\alpha$. If we reject when $p < \alpha$, then we are trying to control our type I error at level $\alpha$.

### Simulation

The following code simulates $n = 100$ observations from the logistic regression model

$$\begin{pmatrix} X_{i1} \\ X_{i2} \end{pmatrix} \overset{iid}{\sim} N\left(\boldsymbol{0}, \begin{pmatrix}1 & 0 \\ 0 & 1 \end{pmatrix} \right)$$

$$Y_i|\mathbf{x}_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2}$$

In this code, $\beta_1 = 0$ and $\beta_2 = 0$ (so $H_0$ is true!). The code then calculates a p-value for the *separate* tests $H_{0,1}: \beta_1 = 0$ and $H_{0,2}: \beta_2 = 0$. Finally, we calculate the Wald test statistic for the joint test of $H_0: \beta_1 = \beta_2 = 0$.


```{r, eval=F}
set.seed(82932)

beta <- c(1, 0, 0)
n <- 100

X <- MASS::mvrnorm(n, mu=c(0, 0), Sigma = matrix(c(1, 0, 0, 1), ncol=2))
X <- cbind(1, X)
p <- exp(as.numeric(X %*% beta))/(1 + exp(as.numeric(X %*% beta)))
y <- rbinom(n, 1, p)
  
m1 <- glm(y ~ X[,2] + X[,3], family = binomial)
summary(m1)$coefficients[2,4] # p-value for testing beta1 = 0
summary(m1)$coefficients[3,4] # p-value for testing beta2 = 0

# joint test
c_mat <- matrix(c(0, 1, 0, 0, 0, 1), nrow = 2, byrow = T)
test_stat <- t(c_mat %*% m1$coefficients) %*% solve(c_mat %*% vcov(m1) %*% t(c_mat)) %*% 
  (c_mat %*% m1$coefficients)

# p-value for Wald test:
pchisq(test_stat, 2, lower.tail=F)
```

### Questions: type I error rate

4. Run the code many times, generating many simulated datasets, and store the three different p-values for each repetition.

5. If we reject $H_0: \beta_1 = \beta_2 = 0$ when the p-value for the joint Wald test is $< 0.05$, verify (with the simulation output) that the type I error rate is approximately 0.05.

6. If we reject $H_0: \beta_1 = \beta_2 = 0$ when $\min\{p_1, p_2\} < 0.05$ (the minimum of the two separate p-values is less than 0.05), verify that the type I error rate is **greater** than 0.05.

7. If we reject $H_0: \beta_1 = \beta_2 = 0$ when $\min\{p_1, p_2\} < 0.025$ (the minimum of the two separate p-values is less than 0.025), verify that the type I error rate is approximately 0.05.

**Moral so far:** If we want to simultaneously test multiple parameters ($\beta_1$ and $\beta_2$) by testing each parameter *separately*, we have to be careful about when we reject $H_0$ in order to control our type I error rate at the desired level!

### Questions: power

Now let's change the simulation so that $H_0$ is *false*. 

8. Copy your code from the previous section, but this time change `beta` in the code so that `beta <- c(1, 0.5, 0.5)`. Leave everything else the same, and re-run the code (many times), storing the three different p-values for each repetition.

9. If we reject $H_0: \beta_1 = \beta_2 = 0$ when the p-value for the joint Wald test is $< 0.05$, what fraction of the time do we (correctly) reject $H_0$?

10. If we reject $H_0: \beta_1 = \beta_2 = 0$ when $\min\{p_1, p_2\} < 0.025$ (the minimum of the two separate p-values is less than 0.025), what fraction of the time do we (correctly) reject $H_0$?

11. Which of the two approaches (testing jointly vs. separately) is more powerful?

### Questions: power, part deux

In questions 8 -- 11, we had $\beta_1 = 0.5$ and $\beta_2 = 0.5$. What if only one of the parameters is truly nonzero?

12. Copy your code from the previous section, but this time change `beta` in the code so that `beta <- c(1, 0, 0.7)`. Leave everything else the same, and re-run the code (many times), storing the three different p-values for each repetition.

13. If we reject $H_0: \beta_1 = \beta_2 = 0$ when the p-value for the joint Wald test is $< 0.05$, what fraction of the time do we (correctly) reject $H_0$?

14. If we reject $H_0: \beta_1 = \beta_2 = 0$ when $\min\{p_1, p_2\} < 0.025$ (the minimum of the two separate p-values is less than 0.025), what fraction of the time do we (correctly) reject $H_0$?

15. Which of the two approaches (testing jointly vs. separately) is more powerful?

**Moral so far:** Testing the parameters $\beta_1, \beta_2$ simultaneously is **not** equivalent to testing them separately!

### Questions: power part 3: return of the jedi

So far, we have simulated our data so that the two explanatory variables $X_{i1}$ and $X_{i2}$ are *independent*. **However**, recall that the interpretation of $\beta_1$ and $\beta_2$ in the model are conditional on having the *other* variable in the model. E.g., the null hypothesis $H_{0,2}: \beta_2 = 0$ is the hypothesis that there is no relationship between $X_{i2}$ and $Y_i$, *after accounting for* $X_{i1}$.

If $X_{i1}$ and $X_{i2}$ are correlated, we can imagine that this might change the results of our hypothesis tests!

16. Copy your code from the previous section, but this time change `beta` in the code so that `beta <- c(1, 0.3, 0.3)` and `Sigma = matrix(c(1, 0.75, 0.75, 1), ncol=2)`. Leave everything else the same, and re-run the code (many times), storing the three different p-values for each repetition.

17. If we reject $H_0: \beta_1 = \beta_2 = 0$ when the p-value for the joint Wald test is $< 0.05$, what fraction of the time do we (correctly) reject $H_0$?

18. If we reject $H_0: \beta_1 = \beta_2 = 0$ when $\min\{p_1, p_2\} < 0.025$ (the minimum of the two separate p-values is less than 0.025), what fraction of the time do we (correctly) reject $H_0$?

19. Which of the two approaches (testing jointly vs. separately) is more powerful?

**Moral:** The **meaning** of the joint test is different than the meaning of combining the separate tests!

# Wald tests: power

Consider the Wald test for a population mean which we discussed in class. Suppose that $X_1,...,X_n$ are a sample from some population with mean $\mu$, and we want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$
Suppose we know the population variance $\sigma^2$. Our test statistic is then

$$Z_n = \frac{\sqrt{n}(\overline{X}_n - \mu_0)}{\sigma}$$

and we reject $H_0$ when $|Z_n| > z_{\alpha/2}$, where $z_{\alpha/2}$ is the upper $\alpha/2$ quantile of a $N(0, 1)$ distribution, and $\alpha$ is our desired Type I error rate. E.g., if $\alpha = 0.05$, then $z_{\alpha/2} \approx 1.96$.

## Power and type I error

*Power* and *type I error* tell us about the *probability* that we reject $H_0$, under different circumstances. Formally, for the Wald test described above,

$$Power(\mu^*) = P(\text{reject } H_0 | \mu = \mu^*)$$
Ideally, we want a *high* probability of rejecting $H_0$, if $H_A$ is true. So, we want $Power(\mu^*)$ to be high for values of $\mu^* \neq \mu_0$. But if $H_0$ is true, we want to control the probability of rejecting (because a rejection would be a false positive!). We call $Power(\mu_0)$ the *type I error rate*. Typically, we want $Power(\mu_0) = \alpha$ (where we get to choose $\alpha$, and common values include $\alpha = 0.01$ or $\alpha = 0.05$).

## Simulations

For the purpose of this activity, suppose that $X_1,...,X_n \overset{iid}{\sim} N(\mu, 1)$. We wish to test

$$H_0: \mu = 0 \hspace{1cm} H_A: \mu \neq 0$$

The following code simulates data with $n = 20$ and $\mu = 0$, calculates the test statistic, and checks whether we reject:


```{r, eval=F}
alpha <- 0.05
n <- 20
mu <- 0
x <- rnorm(n, mu)
z <- sqrt(n)*(mean(x) - 0)/1

# check if reject
abs(z) > qnorm(alpha/2, lower.tail=F)
```

20. Repeat the code above many times, and verify that the fraction of times you reject $H_0$ is approximately 0.05. (This is the *type I error*, i.e. $Power(\mu_0)$)

21. For fixed sample size $n = 20$, estimate $Power(\mu^*)$ for different values of $\mu^*$, and plot $Power(\mu^*)$ against $\mu^*$. How does power (our probability of rejecting) change as we change $\mu^*$?

22. For fixed $\mu^* = 0.2$, estimate $Power(0.2)$ for different values of $n$, and plot $Power(0.2)$ against $n$. How does power (our probability of rejecting) change as we change $n$?

