---
title: "Class activity"
format: html
editor: source
---

# Interim Analysis

Suppose a group of researchers is running an experiment over the course of $n$ months, with a single observation $X_i$ collected each month. From prior studies, the researchers know that
$$X_i \overset{iid}{\sim} N(\mu, 1)$$
They wish to test the hypotheses $H_0: \mu = 0$ vs. $H_A: \mu > 0$. Let
$$Z_k = \frac{1}{\sqrt{k}} \sum \limits_{i=1}^k X_i$$
**Initial plan:** reject $H_0$ if $Z_n > z_{\alpha}$

However, collecting data is expensive, and so they want to save some time and money by looking at the results partway through the experiment. So, the researchers will reject if either $Z_k > c_1$ or $Z_n > c_2$, for some interim point $k$ and some thresholds $c_1, c_2$.

## Questions

First, suppose $n = 30$ and $k = 20$. The researchers decide to use $\alpha = 0.05$ and initially suggest $c_1 = c_2 = z_{\alpha} = 1.645$. 

The code below simulates $X_1,...,X_{30} \overset{iid}{\sim} N(0, 1)$ and calculates $Z_{20}$ and $Z_{30}$.

```{r, eval=F}
c1 <- 1.645
c2 <- 1.645
n <- 30
k <- 20

# simulate data
x <- rnorm(n)

# calculate test statistics
zk <- 1/sqrt(k) * sum(x[1:k])
zn <- 1/sqrt(n) * sum(x)

# check if we reject
(zk > c1) || (zn > c2)
```


1. Run the code. Do you reject $H_0$, using the rule "reject $H_0$ if either $Z_{20} > 1.645$ or $Z_{30} > 1.645$"?

2. Repeat the simulation many times, and approximate the type I error for the researchers' decision rule. Are they controlling type I error at level 0.05?

```{r, include=F}
c1 <- 1.645
c2 <- 1.645
n <- 30
k <- 20

nsim <- 50000
test_res <- rep(NA, nsim)
for(i in 1:nsim){
  # simulate data
  x <- rnorm(n)

  # calculate test statistics
  zk <- 1/sqrt(k) * sum(x[1:k])
  zn <- 1/sqrt(n) * sum(x)

  # check if we reject
  test_res[i] <- (zk > c1) || (zn > c2)
}

mean(test_res)
```


3. Using Bonferroni's method, choose values of $c_1$ and $c_2$ so that you control the type I error at level 0.05. Verify the type I error control in your simulations.

```{r, include=F}
c1 <- 1.96
c2 <- 1.96
n <- 30
k <- 20

nsim <- 1000
test_res <- rep(NA, nsim)
for(i in 1:nsim){
  # simulate data
  x <- rnorm(n)

  # calculate test statistics
  zk <- 1/sqrt(k) * sum(x[1:k])
  zn <- 1/sqrt(n) * sum(x)

  # check if we reject
  test_res[i] <- (zk > c1) || (zn > c2)
}

mean(test_res)
```

Bonferroni's method is conservative -- perhaps we can get a bit more power by finding the exact type I error as a function of the cutoffs $c_1, c_2$. 

For the setting described above, it can be shown that
$$\begin{aligned}
P(Z_{20} > c_1 \cup Z_{30} > c_2) &= P(Z_{20} > c_1) + P(Z_{20} < c_1, Z_{30} > c_2) \\
&= (1 - \Phi(c_1)) + \int \limits_{-\infty}^{c_1} \left(1 - \Phi \left( \sqrt{\frac{30}{10}} \left(c_2 - t\sqrt{\frac{20}{30}} \right) \right) \right) \phi(t) dt
\end{aligned}$$
where $\Phi$ is the standard normal CDF, and $\phi$ is the standard normal pdf.

4. Solving the integral in closed form will be tricky. Instead, let's use R to approximate the integral. Use the `integrate` function in R to calculate the quantity in question 3 when $c_1 = 1.645$ and $c_2 = 1.645$. You may use `pnorm` for $\Phi$ and `dnorm` for $\phi$.

```{r, include=F}
c1 <- 1.645
c2 <- 1.645
integrand <- function(t){
  (1 - pnorm(sqrt(30/10)*(c2 - t*sqrt(20/30))))*dnorm(t)
}
1 - pnorm(c1) + integrate(integrand, lower = -Inf, upper = c1)$value
```


Now we need to figure out thresholds that actually control the type I error rate as desired. One approach is to choose $c_1$ such that $P(Z_{20} > c_1) = \alpha_1$, and then choose $c_2$ such that $P(Z_{20} < c_1, Z_{30} > c_2) = \alpha_2$, with $\alpha_1 + \alpha_2 = \alpha$.

5. To make it simpler, let's do $\alpha_1 = \alpha_2 = 0.025$. Then, $c_1 = 1.96$. By trying different values of $c_2$ in the integral, find a value $c_2$ such that  $P(Z_{20} < 1.96, Z_{30} > c_2) \approx 0.025$.

```{r, include=F}
c1 <- 1.96
c2 <- 1.75
integrand <- function(t){
  (1 - pnorm(sqrt(30/10)*(c2 - t*sqrt(20/30))))*dnorm(t)
}
1 - pnorm(c1) + integrate(integrand, lower = -Inf, upper = c1)$value
```

6. Finally, if you have time: show that

$$\begin{aligned}
P(Z_{20} > c_1 \cup Z_{30} > c_2) &= P(Z_{20} > c_1) + P(Z_{20} < c_1, Z_{30} > c_2) \\
&= (1 - \Phi(c_1)) + \int \limits_{-\infty}^{c_1} \left(1 - \Phi \left( \sqrt{\frac{30}{10}} \left(c_2 - t\sqrt{\frac{20}{30}} \right) \right) \right) \phi(t) dt
\end{aligned}$$

